---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.  

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data
```{r LoadAndPreProcChk,echo=TRUE}
options(scipen = 1, digits = 2)            # numbers > 10^5 in scientific notation, 2 sig figs
library(dplyr,     warn.conflicts = FALSE)
library(stringr,   warn.conflicts = FALSE)
library(ggplot2,   warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(knitr,     warn.conflicts = FALSE)

# Download data if necessary
fnActivity <- "activity.csv"

if ( !file.exists(fnActivity )) {
    message("Data file(s) missing, downloading....")
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", fnActivity)
    message("Unzipping data files...")
    unzip(fnActivity)
}
# Read data
activity <- read.csv(file = "activity.csv",header = TRUE, sep = ",",stringsAsFactors = FALSE)

# mean total no. steps per day
activity.by.day     <- group_by(activity, as.factor(activity$date))
activity.by.day.sum <- summarize(activity.by.day, sum(steps))

# add names to allow further calculations
names(activity.by.day.sum) <- c("Date", "TotalSteps")

# calculate measures of centre
stepsmean   <- mean(  activity.by.day.sum$TotalSteps, na.rm = TRUE)
stepsmedian <- median(activity.by.day.sum$TotalSteps, na.rm = TRUE)

# Prepare data for plotting
activity.by.day.sum$Date <- as.POSIXlt(strptime(activity.by.day.sum$Date, "%Y-%m-%d"))
qplot(x=activity.by.day.sum$Date,y=activity.by.day.sum$TotalSteps,
      geom="histogram",stat="identity", 
      xlab="Date", ylab="Steps per day", main="Total steps per day, excluding missing values")
```

## What is mean total number of steps taken per day?
The mean number of steps taken per day is `r stepsmean`.  
The median number of steps taken per day is `r stepsmedian`.

## What is the average daily activity pattern?
The following code plots the mean number of steps per 5-minute interval across the observation period.  
```{r AvgDailyPatternChk,echo=TRUE}
activity.by.interval <- group_by(activity,as.factor(activity$interval))

activity.by.interval.mean <- summarize(activity.by.interval, mean(steps,na.rm = TRUE))
names(activity.by.interval.mean) <- c("interval", "stepsperinterval")

# Transform interval
activity.by.interval.mean$interval <- str_pad(as.character(activity.by.interval.mean$interval),4,side="left",pad="0")  #convert & left pad with zeroes
activity.by.interval.mean$interval <- as.POSIXct(strptime(activity.by.interval.mean$interval, "%H%M"))   # TODO: contains day of year as well, remove

qplot(x=activity.by.interval.mean$interval, y=activity.by.interval.mean$stepsperinterval, geom="line",  
    scale_x_datetime, minor_breaks = "1 hour", 
    xlab="Interval", ylab="Mean steps per interval", main="Average daily activity pattern")
```

This code determines the maximum number of steps and the interval in which this maximum occurs.
```{r MaxStepsIntervalChk,echo=TRUE}
maxsteps <- max(activity.by.interval.mean$stepsperinterval)
maxint   <- filter(activity.by.interval.mean, stepsperinterval== maxsteps)
```

The maximum number of steps in a 5-minute interval is `r maxsteps`. This occurs between `r hour(maxint$interval)`:`r minute(maxint$interval) - 5` and `r hour(maxint$interval)`:`r minute(maxint$interval)`. 

## Inputting missing values
There are `r nrow(activity) - sum(complete.cases(activity))` missing values in the Activity dataset, calculated from:  
```{r CalcMissingValsChk,echo=TRUE}
countNA <- nrow(activity) - sum(complete.cases(activity))
```
To fill a missing interval value, the strategy is to replace the missing value by the mean value for the same period over the observation period (61 days). For example, if 00:40 to 00:45 is NA on 1 October, replace it with the mean calculated from all days where there is a value; in this case the mean for the period 00:40 to 00:45 is `r activity.by.interval.mean$stepsperinterval[10]`.

```{r FillMissingValsChk,echo=TRUE}
# Create a new dataset
activitycomplete <- activity

# if an interval is NA, fill it in with the mean value, otherwise copy the existing value
activitycomplete$steps <- ifelse(is.na(activity$steps), activity.by.interval.mean$stepsperinterval, activity$steps)
```

The total steps per day, mean and median can now be recalculated as follows:
```{r FilledTotalsChk,echo=TRUE}
# mean total no. steps per day
activitycomplete.by.day     <- group_by(activitycomplete, as.factor(activitycomplete$date))
activitycomplete.by.day.sum <- summarize(activitycomplete.by.day, sum(steps))

# add names to allow further calculations
names(activitycomplete.by.day.sum) <- c("Date", "TotalSteps")

# calculate measures of centre
stepsmeancomplete   <- mean(  activitycomplete.by.day.sum$TotalSteps)  #don't need na.rm here
stepsmediancomplete <- median(activitycomplete.by.day.sum$TotalSteps)

# Prepare data for plotting
activitycomplete.by.day.sum$Date <- as.POSIXlt(strptime(activitycomplete.by.day.sum$Date, "%Y-%m-%d"))
qplot(x=activitycomplete.by.day.sum$Date,y=activitycomplete.by.day.sum$TotalSteps,
      geom="histogram",stat="identity", 
      xlab="Date", ylab="Steps per day - missing values filled", main="Total steps per day, no missing values")
```

The mean number of steps taken per day calculated using the filled-in data is `r stepsmeancomplete`, compared with `r stepsmean` using the original data.  
The median number of steps taken per day calculated using the filled-in data is `r stepsmediancomplete`, compared with `r stepsmedian` using the original data.
The values calculated from the filled-in data are almost identical to those calculated from the original, "gappy" data, indicating that the measures of centre (mean/median) are not greatly changed by filling in missing data. 

## Are there differences in activity patterns between weekdays and weekends?

To examine activity patterns on weekends vs weekdays, we first add a factor indicating "weekday" or "weekend", then change the interval to a
factor so we can group it 
```{r ActivityPatternsChk,echo=TRUE}
activitycomplete$day <- as.factor(ifelse((wday(activitycomplete$date)==1)|(wday(activitycomplete$date)==7),"weekend","weekday"))  
# Now determine the means across two separate observation periods: 2 days for weekend, 5 days for weekday

a <- activitycomplete                    #just duplicate it.
a$date <- as.character(a$date)           #avoid POSIX error in group_by

a$interval <- as.factor(a$interval)               #so we can take the mean
names(a) <- c("steps","date","intervalf","dayf")  #indicate the latter two variables are factors

a.int <- group_by(a, intervalf, dayf)   #group each interval in either weekday or weekend
s     <- summarise(a.int, mean(steps))  #get the mean of each interval; gives two rows for each interval; one for weekday, one for weekend.

names(s) <- c("interval","dayf", "meansteps")  #add names
# Convert to a POSIX format so x axis shows times. The current day is inserted, haven't figure out how to remove this.
s$interval <- str_pad(as.character(s$interval),4,side="left",pad="0")
s$interval <- as.POSIXlt(strptime(s$interval, "%H%M"))
print(ggplot(s, aes(x=interval, y=meansteps)) + geom_line() + facet_grid(dayf ~ .))
```

The patterns show that weekday activity has a higher peak in the morning with less variation during the day, compared with weekends. The peak activities occur around 9am in both cases. The weekend pattern also shows more activity later at night.